---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 10 @ 11:59PM
author: Cordelia Lee UID 105361901
format:
  html:
    theme: cosmo
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
---

Display machine information for reproducibility:

```{r}
#| eval: true
sessionInfo()
```

Load necessary libraries (you can add more as needed).

```{r setup}
library(data.table)
library(lubridate)
library(R.utils)
library(tidyverse)
library(dplyr)
```

MIMIC data location

```{r}
#mimic_path <-"C:\\Users\\Cordelia\\mimic"<- "~/mimic"
mimic_path <- "C:\\Users\\Cordelia\\mimic-iv-1.0"
```

Comment: the link is not working for my laptop because the home setting for R . Dr. Zhou provided this temporary solution so that I can continue to work on this homework.

In this exercise, we use tidyverse (ggplot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2023winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

Display the contents of MIMIC data folder.

```{bash}
ls -l ~/mimic/
```

```{bash}
ls -l ~/mimic/core/
```

```{bash}
ls -l ~/mimic/hosp/

```

```{bash}
ls -l ~/mimic/icu/
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package.

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the `read_csv` in tidyverse.

```{r}
system.time(read.csv(str_c(mimic_path, "/core/admissions.csv.gz")))
# read_csv("~/mimic/core/admissions.csv.gz")
system.time(read_csv(str_c(mimic_path, "/core/admissions.csv.gz")))
system.time(fread(str_c(mimic_path, "/core/admissions.csv.gz")))

```

Ans: the one that appears to be the fastest is the read.csv function. The read_csv one seemed to be the slowest. There is a difference between the default and the one in tidyverse. read_csv noticed some issues with the column.

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are

```{bash}
zcat ~/mimic/icu/icustays.csv.gz | head -10

```

1.  Import `icustatys.csv.gz` as a tibble `icustays_tble`.

    ```{r}
    icu<-(read_csv(str_c(mimic_path, "/icu/icustays.csv.gz")))
    icustays_tble <- as.tibble(icu)
    ```

2.  How many unique `subject_id`? Can a `subject_id` have multiple ICU stays?

    ```{r}

    n_distinct (icustays_tble$subject_id)

    ```

3.  Summarize the number of ICU stays per `subject_id` by graphs.

    ```{r}
    icu_count<-icustays_tble %>% 
      group_by(subject_id)%>% 
      summarise(COUNT = n())
    hist(x=icu_count$COUNT, main="ICU stays of a patient")
    ```

4.  For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`. (Hint: `slice_min` and `slice_max` may take long. Think alternative ways to achieve the same function.)

    ```{r}
    icustays_tble <- icustays_tble %>%
      group_by(subject_id) %>%
      arrange(intime, .by_group = TRUE) %>%
      slice_head(n=1)%>%
      ungroup()%>%
      print(width = Inf)
    #distinct would also work
    #   arrange(subject_id, intime) %>%
    #   distinct(subject_id, .keep_all = TRUE) %>%
    ```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are

```{bash}
zcat ~/mimic/core/admissions.csv.gz| head -10

```

1.  Import `admissions.csv.gz` as a tibble `admissions_tble`.

    ```{r}
    admin<-(read_csv(str_c(mimic_path, "/core/admissions.csv.gz")))
    admissions_tble <- as.tibble(admin)
    ```

2.  Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadmi_id`.

    ```{r}
    # admissions_tble <-admissions_tble%>%
    #   semi_join(icustays_tble, by=c('subject_id','hadmi_id'))%>%
    #   print(width=Inf)
    semi_join(x = admissions_tble, y = icustays_tble, by = c("subject_id", "hadm_id"))
    ```

3.  Summarize the following variables by graphics.

    -   admission year\

    ```{r}
    admissions_tble %>%
      ggplot(aes(x=year(admittime)))+
      geom_bar()+
      labs(x="admission year")

    ```

    -   admission month\

    ```{r}
    admissions_tble %>%
      ggplot(aes(x=months.Date(admittime)))+
      geom_bar()+
      labs(x="admission month")
    ```

    -   admission month day

    ```{r}
    admissions_tble %>%
      ggplot(aes(x=mday(admittime)))+
      geom_bar()+
      labs(x="admission month day")
    ```

    -   admission week day\

    ```{r}
    admissions_tble %>%
      ggplot(aes(x=weekdays.Date(admittime)))+
      geom_bar()+
      labs(x="admission week day")
    ```

    -   admission hour (anything unusual?)\
        There are a lot of patients admitted at midnight, which are likely results of randomly inputting because it was not recorded.

    ```{r}
    admissions_tble %>%
      ggplot(aes(x=hour(admittime)))+
      geom_bar()+
      labs(x="admission hour")
    ```

    -   admission minute (anything unusual?)\
        There are also a lot of patients admitted at 0 hours, which may be due to similar reasons as above with admission hour.

    ```{r}
    admissions_tble %>%
      ggplot(aes(x=minute(admittime)))+
      geom_bar()+
      labs(x="admission minute")
    ```

    -   length of hospital stay (anything unusual?)\
        There are patients with 0 hours of hospital stay.

    ```{r}
    admissions_tble %>%
      ggplot(aes(x=difftime(dischtime,admittime, unit="hours")))+
      geom_bar()+
      xlim(0,1800)+
      ylim(0,130)+
      labs(x="length of hispital stay by hour")
    ```

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are

```{bash}
zcat ~/mimic/core/patients.csv.gz| head -10
```

1.  Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

    ```{r}
    patients_tble<-
      read_csv(str_c(mimic_path, "/core/patients.csv.gz")) %>%
      semi_join (icustays_tble, by= "subject_id") %>%
      print (width=Inf)
    ```

2.  Summarize variables `gender` and `anchor_age`, and explain any patterns you see.

    ```{r}
    patients_tble %>%
      ggplot(aes(x=gender))+
      geom_bar()+
      labs(x="Summary of gender")
    ```

    There are more males than females in the patients_tble.

    ```{r}
    patients_tble %>%
      ggplot(aes(x=anchor_age))+
      geom_bar()+
      labs(x="Summary of anchor_age")
    ```

There is an unusual amount of people who are 91 years old and no one for 90 and below 18. It is likely that the participants below 18 were rare and categorized as 18. Patients above 90 are recorded as 91.

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are

```{bash}
zcat ~/mimic//hosp/labevents.csv.gz| head -10
```

`d_labitems.csv.gz` is the dictionary of lab measurements.

```{bash}
zcat ~/mimic//hosp/d_labitems.csv.gz| head -10
```

1.  Find how many rows are in `labevents.csv.gz`.

    ```{bash}
    zcat ~/mimic/hosp/labevents.csv.gz | tail -n +2 | wc -l
    ```

Ans: There are 122103667 rows excluding the first line of column names in the data frame.

2.  We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), and glucose (50931). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`.

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer 5-10 minutes to ingest this file. If your computer really has trouble importing `labevents.csv.gz`, you can import from the reduced data file `labevents_filtered_itemid.csv.gz`.

```{r}
dlabitems_tble<-read_csv(str_c(mimic_path,"/hosp/d_labitems.csv.gz"))%>%
    filter(itemid %in% c(
      50912,
      50971,
      50983,
      50902,
      50882,
      51221,
      51301,
      50931
      ))%>%
    print(width=Inf)
```

```{r}
    labevents_tble<-
      read_csv(str_c(mimic_path, "/hosp/labevents_filtered_itemid.csv.gz"), 
               col_select = c(subject_id,itemid,charttime,valuenum),
               col_types = cols_only(subject_id=col_double(),
                                     itemid=col_double(),
                                     charttime=col_datetime(),
                                     valuenum=col_double())
               )%>%
      semi_join(icustays_tble,by=c("subject_id"))%>%
      filter(itemid %in% dlabitems_tble$itemid)%>%
      print(width=Inf)
```

3.  Further restrict `labevents_tble` to the first lab measurement during the ICU stay.

```{r}
    labevents_tble %>%
      #only keep items during the frist ICU stay
      left_join(select(icustays_tble, subject_id, intime, outtime),
     by = c( "subject_id")) %>% 
      filter(charttime >= intime & charttime <= outtime) %>%
       group_by(subject_id, itemid) %>%
      arrange (charttime, .by_group= TRUE) %>%
       slice_head(n=1) %>%
       ungroup() %>%
      # important to take out before pivot_wider or else multiple rows for an itemid woll be produced
       select(-intime,-outtime,-charttime) %>%
       pivot_wider(names_from = itemid, values_from=valuenum) %>%
      rename_at(
        vars(as.character(dlabitems_tble$itemid)),
             ~str_to_lower(dlabitems_tble$label)
      ) %>%
       print (width=Inf) 
    
```

4.  Summarize the lab measurements by appropriate numerics and graphics.

    ```{r}
    labevents_tble %>%
      ggplot(aes(x=year(charttime)))+
      geom_bar()+
      labs(x="charttime year")
    ```

```{r}
labevents_tble %>%
  ggplot(aes(x=hour(charttime)))+
  geom_bar()+
  labs(x="charttime hour")

```

```{r}
labevents_tble %>%
  ggplot(aes(x=year(charttime)))+
  geom_bar()+
  labs(x="charttime minue")

```

```{r}
labevents_tble %>%
  ggplot(aes(x=mday(charttime)))+
  geom_bar()+
  labs(x="charttime month day")
```

```{r}
labevents_tble %>%
  ggplot(aes(x=weekdays.Date(charttime)))+
  geom_bar()+
  labs(x="charttime weekdays")
```

Ans: From the graphs above, we can see that the mode for chart time is around 6 hours. The closer to present years also had fewer chart times than the past. Also, chart times seems to be evenly distributed through the month and week, unlike the ICU stays date that had peaks.

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient's information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are

```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/chartevents.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```

`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`.

```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/d_items.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```

1.  We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

    Hint: `chartevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer \>15 minutes to ingest this file. If your computer really has trouble importing `labevents.csv.gz`, you can import from the reduced data file `chartevents_filtered_itemid.csv.gz`.

2.  Further restrict `chartevents_tble` to the first vital measurement during the ICU stay.

3.  Summarize these vital measurements by appropriate numerics and graphics.

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are the first ICU stay of each unique adult (age at admission \> 18) and columns contain at least following variables

-   all variables in `icustays.csv.gz`\
-   all variables in `admission.csv.gz`\
-   all variables in `patients.csv.gz`\
-   first lab measurements during ICU stay\
-   first vital measurements during ICU stay
-   an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

-   `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

-   `thirty_day_mort` vs first lab measurements

-   `thirty_day_mort` vs first vital measurements

-   `thirty_day_mort` vs first ICU unit
