---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 10 @ 11:59PM
author: Cordelia Lee UID 105361901
format:
  html:
    theme: cosmo
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
---

Display machine information for reproducibility:

```{r}
#| eval: true
sessionInfo()
```

Load necessary libraries (you can add more as needed).

```{r setup}
library(data.table)
library(lubridate)
library(R.utils)
library(tidyverse)
```

MIMIC data location

```{r}
#mimic_path <-"C:\\Users\\Cordelia\\mimic"<- "~/mimic"
mimic_path <- "C:\\Users\\Cordelia\\mimic-iv-1.0"
```

Comment: the link is not working for my laptop because the home setting for R . Dr. Zhou provided this temporary solution so that I can continue to work on this homework.

In this exercise, we use tidyverse (ggplot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2023winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

Display the contents of MIMIC data folder.

```{bash}
ls -l ~/mimic/
```

```{bash}
ls -l ~/mimic/core/
```

```{bash}
ls -l ~/mimic/hosp/

```

```{bash}
ls -l ~/mimic/icu/
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package.

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the `read_csv` in tidyverse.

```{r}
system.time(read.csv(str_c(mimic_path, "/core/admissions.csv.gz")))
# read_csv("~/mimic/core/admissions.csv.gz")
system.time(read_csv(str_c(mimic_path, "/core/admissions.csv.gz")))
system.time(fread(str_c(mimic_path, "/core/admissions.csv.gz")))

```

Ans: the one that appears to be the fastest is the read.csv function. The read_csv one seemed to be the slowest. There is a difference between the default and the one in tidyverse. read_csv noticed some issues with the column.

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are

```{bash}
zcat ~/mimic/icu/icustays.csv.gz | head -10

```

1.  Import `icustatys.csv.gz` as a tibble `icustays_tble`.

    ```{r}
    icu<-(read_csv(str_c(mimic_path, "/icu/icustays.csv.gz")))
    icustays_tble <- as.tibble(icu)
    ```

2.  How many unique `subject_id`? Can a `subject_id` have multiple ICU stays?

    ```{r}

    icu_sub_ID<-n_distinct (icustays_tble$subject_id)
    icu_sub_ID
    ```

3.  Summarize the number of ICU stays per `subject_id` by graphs.

    ```{r}
    #ggplot(data=icustays_tble$intime)
    #ggplot2::fortify(icustays_tble)

    #NEEDS MORE WORK
    ```

4.  For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`. (Hint: `slice_min` and `slice_max` may take long. Think alternative ways to achieve the same function.)

    ```{r}
    icustays_tble <- icustays_tble %>%
      arrange(subject_id, intime) %>%
      distinct(subject_id, .keep_all = TRUE) %>%
      print(width = Inf)
    ```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are

```{bash}
zcat ~/mimic/core/admissions.csv.gz| head -10

```

1.  Import `admissions.csv.gz` as a tibble `admissions_tble`.

    ```{r}
    admin<-(read_csv(str_c(mimic_path, "/core/admissions.csv.gz")))
    admissions_tble <- as.tibble(admin)
    ```

2.  Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadmi_id`.

3.  Summarize the following variables by graphics.

    -   admission year\
    -   admission month\
    -   admission month day\
    -   admission week day\
    -   admission hour (anything unusual?)\
    -   admission minute (anything unusual?)\
    -   length of hospital stay (anything unusual?)

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are

```{bash}
zcat ~/mimic/core/patients.csv.gz| head -10
```

1.  Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

2.  Summarize variables `gender` and `anchor_age`, and explain any patterns you see.

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are

```{bash}
zcat ~/mimic//hosp/labevents.csv.gz| head -10
```

`d_labitems.csv.gz` is the dictionary of lab measurements.

```{bash}
zcat ~/mimic//hosp/d_labitems.csv.gz| head -10
```

1.  Find how many rows are in `labevents.csv.gz`.

    ```{r}
    #labevents_tble<-read_csv(str_c(mimic_path, "/hosp/labevents.csv.gz"))
    #nrow(labevents_tble)
    ```

2.  We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), and glucose (50931). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`.

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer 5-10 minutes to ingest this file. If your computer really has trouble importing `labevents.csv.gz`, you can import from the reduced data file `labevents_filtered_itemid.csv.gz`.

3.  Further restrict `labevents_tble` to the first lab measurement during the ICU stay.

4.  Summarize the lab measurements by appropriate numerics and graphics.

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient's information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are

```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/chartevents.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```

`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`.

```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/d_items.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```

1.  We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

    Hint: `chartevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer \>15 minutes to ingest this file. If your computer really has trouble importing `labevents.csv.gz`, you can import from the reduced data file `chartevents_filtered_itemid.csv.gz`.

2.  Further restrict `chartevents_tble` to the first vital measurement during the ICU stay.

3.  Summarize these vital measurements by appropriate numerics and graphics.

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are the first ICU stay of each unique adult (age at admission \> 18) and columns contain at least following variables

-   all variables in `icustays.csv.gz`\
-   all variables in `admission.csv.gz`\
-   all variables in `patients.csv.gz`\
-   first lab measurements during ICU stay\
-   first vital measurements during ICU stay
-   an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

-   `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

-   `thirty_day_mort` vs first lab measurements

-   `thirty_day_mort` vs first vital measurements

-   `thirty_day_mort` vs first ICU unit
