---
title: "Biostat 203B Homework 4"
subtitle: Due Mar 24 @ 11:59PM
author: Cordelia Lee UID: 105361901
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
---

Display machine information:
```{r}
#| eval: true

sessionInfo()
```
Load database libraries and the tidyverse frontend:
```{r}
#| eval: true

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(tidymodels))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(glmnet))
```

## Predicting 30-day mortality

Using the ICU cohort `icu_cohort.rds` you built in Homework 3, develop at least three analytic approaches for predicting the 30-day mortality of patients admitted to ICU using demographic information (gender, age, marital status, ethnicity), first lab measurements during ICU stay, and first vital measurements during ICU stay. For example, you can use (1) logistic regression with elastic net (lasso + ridge) penalty (e.g., glmnet or keras package), (2) random forest, (3) boosting, and (4) support vector machines, or (5) MLP neural network (keras package)

1. Partition data into 50% training set and 50% test set. Stratify partitioning according the 30-day mortality status.

```{r}

  icu_cohort <- read_rds("mimiciv_shiny/icu_cohort.rds")
set.seed(203)
#try smaller data set first 
# temp <- icu_cohort [1:100, ]
# icu_cohort <-temp
data_split <- initial_split(
  icu_cohort, 
  # stratify by 30 day mortality
  strata = "thirty_day_mort", 
  prop = 0.5
  )
data_split

```


2. Train and tune the models using the training set.
```{r}

 
```


3. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each model.

```{r}
# might do random forest, XGboost (tuning might need to pass n-jobs. 
# will take a long time), SVM need to standardize  
# to speed up, don't do large epochs (do 5) and don't tune too much if MLP 
# MLP do 2 4 8 or 12 for hidden units. don't tune drop out. 
# logistic regression

```

## Logistic Regression 
1. Partition data into 50% training set and 50% test set. Stratify partitioning according the 30-day mortality status.
```{r}
library(GGally)
library(gtsummary)
library(ranger)
library(tidyverse)
library(tidymodels)
# load the data "icu_cohort"
icu_cohort <- read_rds("mimiciv_shiny/icu_cohort.rds")%>% 
# only keep the variables wanted. discarded those with a lot NAs
select ("gender", "marital_status", "ethnicity",
        "thirty_day_mort", "item_50912", "item_50971", "item_50912",
        "item_50971", "item_50983", "item_50902", "item_50882", "item_51221",
        "item_51301", "item_50931", "item_220045", "item_220181", 
        "item_220179", "item_223761", "item_220210")

set.seed(203)

#splitting data 
data_split <- initial_split(
  icu_cohort, 
  # stratify by thirty day mortality
  strata = "thirty_day_mort", 
  prop = 0.5
  )
data_split

icu_cohort_other <- training(data_split)
dim(icu_cohort_other)

icu_cohort_test <- testing(data_split)
dim(icu_cohort_test)
```
2. Train and tune the models using the training set.
```{r}
logit_recipe <- 
  recipe(
    thirty_day_mort ~ ., 
    data = icu_cohort
  ) %>%
  # ignore missing data and impute because sample size is large
  # only impute the variables of interest
   step_impute_mode(marital_status, ethnicity, gender) %>%
   step_impute_mean(item_50912, item_50971, item_50912,
       item_50971,
     item_50983,
      item_50902,
      item_50882,
      item_51221,
      item_51301,
      item_50931,
     item_220045,
      item_220181,
      item_220179,
      item_223761,
      item_220210)%>%
  # create traditional dummy variables
  step_dummy(all_nominal_predictors()) %>%
  # zero-variance filter
  step_zv(all_numeric_predictors()) %>% 
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) %>%
  # estimate the means and standard deviations
  prep(training = icu_cohort_other, retain = TRUE)
logit_recipe
```
Model 
```{r}
logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) %>% 
  set_engine("glmnet", standardize = FALSE)
logit_mod
```
Workflow
```{r}
logit_wf <- workflow() %>%
  add_recipe(logit_recipe) %>%
  add_model(logit_mod)
logit_wf
```
Tuning grid 
```{r}
param_grid <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(100, 5)
  )
param_grid
```
Cross-validation
```{r}
set.seed(203)

folds <- vfold_cv(icu_cohort_other, v = 5)
folds
```
fit CV
```{r}
#str(icu_cohort)
logit_fit <- logit_wf %>%
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )

```

```{r}
logit_fit
```
```{r}
logit_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "roc_auc") %>%
  ggplot(mapping = aes(x = penalty, y = mean, color = mixture)) +
  geom_point() +
  labs(x = "Penalty", y = "CV AUC") +
  scale_x_log10()
```

```{r}
logit_fit %>%
  show_best("roc_auc")
```


```{r}
best_logit <- logit_fit %>%
  select_best("roc_auc")
best_logit
```


##Random Forest 
```{r}

```


